---
title: "Wisconsin Shared Revenue is Redistributive"
author: "Michael J Zenz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    page-layout: full
editor: visual
---

## County-level Distribution of Shared Revenue

```{r, setup}
#| echo: false
#| warning: false
#| message: false


##This R scrip uploads shared revenue data, processes, and provides analyses

library(tidyverse)
library(rvest)

###Process shared revenue data input

#Read in .csv data with shared revenue and income/sales tax data by county
#This was provided by the WI Legislative Fiscal Bureau at the request of Rep. Francesca Hong 
shared.revenue.data.unprocessed <-  read_csv("Shared_Rev_Data.csv")

#Converts values given in milions.
shared.revenue.data <- shared.revenue.data.unprocessed %>%
                        mutate(`Net Individual Income Tax` = `Net Individual Income Tax` * 1000000, 
                               `State Sales Tax Collections` =  `State Sales Tax Collections`  * 1000000)

#Scrapes county population and area data from wikipedia table
wiki.county.table <- rvest::read_html("https://en.wikipedia.org/wiki/List_of_counties_in_Wisconsin") %>%
            rvest::html_node(., ".wikitable") %>%
     rvest::html_table(., fill = TRUE)

#Removes "county" from county names and renames variables.
county.pop.area <- wiki.county.table %>%
                    mutate(County = substr(County, 1,str_length(County)-7)) %>%
                  select(County, Population = `Population[2]`, Area = `Area[2]`)

#Removes commas from population values
county.pop.area$Population <- as.numeric(gsub("[\\,]","",county.pop.area$Population))

#Scrapes WI 2020 presidential election values from wikipedia
wiki.trump.vote.table <- rvest::read_html("https://en.wikipedia.org/wiki/2020_United_States_presidential_election_in_Wisconsin") %>%
  html_nodes(.,"table")
wiki.trump.vote.table <-html_table(wiki.trump.vote.table[35], fill = TRUE)[[1]] 

#Keep Trump 2020 vote share.
trump.vote.prop.table <- wiki.trump.vote.table[-1,c(1,4)] %>%
                      rename(Trump_2020_Prop = Trump) %>%
                      mutate(Trump_2020_Prop = as.numeric(substr(Trump_2020_Prop, 1, 
                                                      str_length(Trump_2020_Prop)-1))/100)

#Process data and make further calculations.
data <- shared.revenue.data %>%
        left_join(county.pop.area, by = "County") %>%
        left_join(trump.vote.prop.table, by = "County") %>%
        mutate(`Area (sq mi)` = as.numeric(str_replace_all(str_extract(Area, "\\d+[\\d,.]*"), ",", "")) ,
               `Density (pop sq mi)` = Population/`Area (sq mi)`,
                `Combined Tax` = `Net Individual Income Tax` + `State Sales Tax Collections`,
               `Tax Prop Shared Revenue` = (`Combined Tax`/sum(`Combined Tax`)) * sum(`Total Shared Revenue`),
               `Total Shared Revenue Subsidy` = `Total Shared Revenue` - `Tax Prop Shared Revenue`,
               `Per capita Shared Revenue Subsidy` =  `Total Shared Revenue Subsidy` / Population,
               `Per capita Income Tax` = `Net Individual Income Tax`/Population,
               `Per capita Combined Tax` = `Combined Tax`/Population,
               `2020 Trump Majority` = Trump_2020_Prop > .5 
               )

rm(county.pop.area, shared.revenue.data, shared.revenue.data.unprocessed, trump.vote.prop.table,
   wiki.county.table, wiki.trump.vote.table)



#Linear Regression model for population and amount of shared revenue.
income.lm <- lm(`Per capita Shared Revenue Subsidy` ~ log(`Per capita Income Tax`), data = data)
#summary(income.lm)


population.lm <- lm(`Total Shared Revenue` ~ Population, data)
#summary(population.lm)
population.lm.log <-  lm(log(`Total Shared Revenue`) ~ log(Population), data)
#summary(population.lm.log)

intercept <- coef(population.lm.log)[1]
slope <- coef(population.lm.log)[2]

```

During the spring and summer of 2023 leaders in the Wisconsin legislature along with Governor Evers proposed changes to county and municipal aid, which is a small part of the total state tax proceeds shared with local governments.[^1] This "shared revenue" is collected primarily through state income and sales taxes and are redistributed to county and municipal governments to help them fund the various services that are typically provided by local rather than state government.

[^1]: For information about state aid see a recent [WI Legislative Fiscal Bureau Report](https://docs.legis.wisconsin.gov/misc/lfb/misc/118_distributions_of_existing_and_supplemental_county_and_municipal_aid_7_10_23.pdf) on changes to state aid to counties. However this is a small fraction of all shared revenue. For instance, state aid to Dane County governments was set to increase from around \$1M to \$5M in 2024, whereas total shared revenue in 2020-21 was around \$250M.

Despite [a complex formula](https://docs.legis.wisconsin.gov/misc/lfb/informational_papers/january_2017/0018_shared_revenue_program_informational_paper_18.pdf) for calculating the level of shared revenue, **population largely determines the level of shared revenue that each county receives**.

```{r, population_graph}
#| echo: false
#| warning: false


ggplot(data = data, aes(x = log(Population), y = log(`Total Shared Revenue`))) + 
            geom_point(aes(color = `2020 Trump Majority`)) +
  scale_colour_manual(values = c("blue", "red")) + 
  geom_abline(slope = slope, intercept = intercept) + 
  labs(title = "WI County (and municipal) Shared Revenue by Population",
       x = "Population (log)", y = "Shared Revenue in $ (log)") +
theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x =element_blank() , 
        axis.text.y = element_blank() , 
        axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank())



```

Individuals are given roughly equal allocations regardless of where they live, but this can result in local governments receiving proportionally less money in shared revenue than their residents pay in. Richer countries tend to receive less than they pay in, and poorer counties more.

::: {layout-ncol="2"}
```{r, shared_revenue_subsidy, fig.height = 10 , fig.width = 6}
#| echo: false
#| warning: false

ggplot(data = data, aes(y = reorder(County,`Per capita Combined Tax`) , x = `Per capita Shared Revenue Subsidy`)) +
  scale_fill_manual(values = c("blue", "red")) + 
  geom_bar(aes(fill = `2020 Trump Majority`), stat = "identity") +
  theme_minimal() +
labs(title = "Per Capita Shared Revenue Subsidy",
     subtitle = "Ordered by per-capita tax procedes (highest to lowest)",
       x = "Per Capita $", y = "") +     
  theme(legend.position="bottom")
  
    
```

```{r, Change_InFunding, fig.height = 10 , fig.width = 6}
#| echo: false
#| warning: false

ggplot(data = data, aes(y = reorder(County,`Per capita Combined Tax`) , x = ( (`Total Shared Revenue` - `Total Shared Revenue Subsidy`) - `Total Shared Revenue`)/`Total Shared Revenue`)) +
  scale_fill_manual(values = c("blue", "red")) + 
   scale_x_continuous(labels = scales::percent) +
  geom_bar(aes(fill = `2020 Trump Majority`), stat = "identity") +
  theme_minimal() +
labs(title = "% change in Shrd. Rev. if Based on Taxes Only",
     subtitle = "Ordered by per-capita tax procedes (highest to lowest)",
       x = "% Change in Shared Revenue", y = "") +     
  theme(legend.position="bottom")
  
    
```
:::

Because of the patterns of wealth in Wisconsin, tax proceeds from richer (primarily suburban) counties generally flow to poorer ones. Additionally, less dense counties receive less per-capita shared revenue than denser ones, with the exception of Milwaukee County.

```{r, density}




#linear model of density and per capita shared revenue, with individual model per quartile.
density.percapita.combinedtax.quartile1 <-  data %>% 
  mutate(`Tax Quartile` = ntile(`Per capita Combined Tax`, 4)) %>%
  filter(`Tax Quartile` == 1) %>%
lm(log((`Total Shared Revenue`)/Population) ~ log(`Density (pop sq mi)`), data = .)
#summary(density.percapita.combinedtax.quartile1)

intercept.slope.df <- data.frame(quartile = 1, 
                                 intercept = as.numeric(coef(density.percapita.combinedtax.quartile1)[1]),
                                 slope = as.numeric(coef(density.percapita.combinedtax.quartile1)[2]))

intercept.1 <- as.numeric(coef(density.percapita.combinedtax.quartile1)[1])
slope.1 <- as.numeric(coef(density.percapita.combinedtax.quartile1)[2])

density.percapita.combinedtax.quartile2 <-  data %>% 
  mutate(`Tax Quartile` = ntile(`Per capita Combined Tax`, 4)) %>%
  filter(`Tax Quartile` == 2) %>%
  lm(log((`Total Shared Revenue`)/Population) ~ log(`Density (pop sq mi)`), data = .)
#summary(density.percapita.combinedtax.quartile2)

intercept.slope.df <- rbind(intercept.slope.df,data.frame(quartile = 2, 
                                   intercept = as.numeric(coef(density.percapita.combinedtax.quartile2)[1]),
                                   slope = as.numeric(coef(density.percapita.combinedtax.quartile2)[2])))

intercept.2 <- coef(density.percapita.combinedtax.quartile2)[1]
slope.2 <- coef(density.percapita.combinedtax.quartile2)[2]

density.percapita.combinedtax.quartile3 <-  data %>% 
  mutate(`Tax Quartile` = ntile(`Per capita Combined Tax`, 4)) %>%
  filter(`Tax Quartile` == 3) %>%
  lm(log((`Total Shared Revenue`)/Population) ~ log(`Density (pop sq mi)`), data = .)
#summary(density.percapita.combinedtax.quartile3)


intercept.slope.df <- rbind(intercept.slope.df,data.frame(quartile = 3, 
                                    intercept = as.numeric(coef(density.percapita.combinedtax.quartile3)[1]),
                                     slope = as.numeric(coef(density.percapita.combinedtax.quartile3)[2])))

intercept.3 <- coef(density.percapita.combinedtax.quartile3)[1]
slope.3 <- coef(density.percapita.combinedtax.quartile3)[2]

density.percapita.combinedtax.quartile4 <-  data %>% 
  mutate(`Tax Quartile` = ntile(`Per capita Combined Tax`, 4)) %>%
  filter(`Tax Quartile` == 4) %>%
  lm(log((`Total Shared Revenue`)/Population) ~ log(`Density (pop sq mi)`), data = .)
#summary(density.percapita.combinedtax.quartile4)


intercept.slope.df <- rbind(intercept.slope.df,data.frame(quartile = 4, 
                                  intercept = as.numeric(coef(density.percapita.combinedtax.quartile4)[1]),
                                slope = as.numeric(coef(density.percapita.combinedtax.quartile4)[2])))
intercept.4 <- coef(density.percapita.combinedtax.quartile4)[1]
slope.4 <- coef(density.percapita.combinedtax.quartile4)[2]

data %>% mutate(quartile = ntile(`Per capita Combined Tax`, 4)) %>%
inner_join(intercept.slope.df, by = "quartile") %>%
mutate(quartile.text = ifelse(quartile > 1, "25% Below Median Tax","Bottom 25% Tax"),
       quartile.text = ifelse(quartile > 2, "25% Above Median Tax", quartile.text),
       quartile.text = ifelse(quartile > 3, "Top 25% Tax", quartile.text),
       quartile.text = factor(quartile.text, levels = c("Bottom 25% Tax","25% Below Median Tax",
                                                        "25% Above Median Tax", "Top 25% Tax"))) %>%
ggplot(data = ., aes(x = log(`Density (pop sq mi)`), y = log((`Total Shared Revenue`)/Population))) +
  geom_point(aes(color = `2020 Trump Majority`)) +
  scale_color_manual(values = c("blue", "red")) + 
  theme_minimal() +
labs(title = "Per Capita Shared Revenue by Density and Per Capita Tax Proceeds (Panes)",
       x = "Population Density (log)", y = "Per Capita Shared Revenue (log)") +     
  theme(legend.position="bottom")+
  facet_grid(cols = vars(quartile.text)) + 
  geom_abline(aes(slope = slope, intercept = intercept))  +
    theme(axis.text.x = element_blank(),
        axis.ticks.x =element_blank() , 
        axis.text.y = element_blank() , 
        axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank())


```

## Should Shared Revenue Redistribute?

The current shared revenue scheme is quite effective at redistributing money from richer counties to poorer ones. The more foundational question is whether this is something state government should do.
