---
title: "Wisconsin Shared Revenue and Wealth Redistribution"
author: "Michael J Zenz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    page-layout: full
editor: visual
---

## The Distribution of Shared Revenue

```{r, setup}
#| echo: false
#| warning: false
#| message: false


##This R scrip uploads shared revenue data, processes, and provides analyses

library(tidyverse)
library(rvest)

###Process shared revenue data input

#Read in .csv data with shared revenue and income/sales tax data by county
#This was provided by the WI Legislative Fiscal Bureau at the request of Rep. Francesca Hong 
shared.revenue.data.unprocessed <-  read_csv("Shared_Rev_Data.csv")

#Converts values given in milions.
shared.revenue.data <- shared.revenue.data.unprocessed %>%
                        mutate(`Net Individual Income Tax` = `Net Individual Income Tax` * 1000000, 
                               `State Sales Tax Collections` =  `State Sales Tax Collections`  * 1000000)

#Scrapes county population and area data from wikipedia table
wiki.county.table <- rvest::read_html("https://en.wikipedia.org/wiki/List_of_counties_in_Wisconsin") %>%
            rvest::html_node(., ".wikitable") %>%
     rvest::html_table(., fill = TRUE)

#Removes "county" from county names and renames variables.
county.pop.area <- wiki.county.table %>%
                    mutate(County = substr(County, 1,str_length(County)-7)) %>%
                  select(County, Population = `Population[2]`, Area = `Area[2]`)

#Removes commas from population values
county.pop.area$Population <- as.numeric(gsub("[\\,]","",county.pop.area$Population))

#Scrapes WI 2020 presidential election values from wikipedia
wiki.trump.vote.table <- rvest::read_html("https://en.wikipedia.org/wiki/2020_United_States_presidential_election_in_Wisconsin") %>%
  html_nodes(.,"table")
wiki.trump.vote.table <-html_table(wiki.trump.vote.table[35], fill = TRUE)[[1]] 

#Keep Trump 2020 vote share.
trump.vote.prop.table <- wiki.trump.vote.table[-1,c(1,4)] %>%
                      rename(Trump_2020_Prop = Trump) %>%
                      mutate(Trump_2020_Prop = as.numeric(substr(Trump_2020_Prop, 1, 
                                                      str_length(Trump_2020_Prop)-1))/100)

#Process data and make further calculations.
data <- shared.revenue.data %>%
        left_join(county.pop.area, by = "County") %>%
        left_join(trump.vote.prop.table, by = "County") %>%
        mutate(`Area (sq mi)` = as.numeric(str_replace_all(str_extract(Area, "\\d+[\\d,.]*"), ",", "")) ,
               `Density (pop sq mi)` = Population/`Area (sq mi)`,
                `Combined Tax` = `Net Individual Income Tax` + `State Sales Tax Collections`,
               `Tax Prop Shared Revenue` = (`Combined Tax`/sum(`Combined Tax`)) * sum(`Total Shared Revenue`),
               `Total Shared Revenue Subsidy` = `Total Shared Revenue` - `Tax Prop Shared Revenue`,
               `Per capita Shared Revenue Subsidy` =  `Total Shared Revenue Subsidy` / Population,
               `Per capita Income Tax` = `Net Individual Income Tax`/Population,
               `Per capita Combined Tax` = `Combined Tax`/Population,
               `2020 Trump Majority` = Trump_2020_Prop > .5 
               )

rm(county.pop.area, shared.revenue.data, shared.revenue.data.unprocessed, trump.vote.prop.table,
   wiki.county.table, wiki.trump.vote.table)



#Linear Regression model for population and amount of shared revenue.
income.lm <- lm(`Per capita Shared Revenue Subsidy` ~ log(`Per capita Income Tax`), data = data)
#summary(income.lm)


population.lm <- lm(`Total Shared Revenue` ~ Population, data)
#summary(population.lm)
population.lm.log <-  lm(log(`Total Shared Revenue`) ~ log(Population), data)
#summary(population.lm.log)

intercept <- coef(population.lm.log)[1]
slope <- coef(population.lm.log)[2]

```

During the spring and summer of 2023 leaders in the Wisconsin legislature along with Governor Evers proposed changes to county and municipal aid, which is a small part of the total state tax proceeds shared with local governments.[^1] This "shared revenue" is collected primarily through state income and sales taxes and are redistributed to county and municipal governments to help them fund the various services that are typically provided by local rather than state government.

[^1]: For information about state aid see a recent [WI Legislative Fiscal Bureau Report](https://docs.legis.wisconsin.gov/misc/lfb/misc/118_distributions_of_existing_and_supplemental_county_and_municipal_aid_7_10_23.pdf) on changes to state aid to counties. However this is a small fraction of all shared revenue. For instance, state aid to Dane County governments was set to increase from around \$1M to \$5M in 2024, whereas total shared revenue in 2020-21 was around \$250M.

Despite [a complex formula](https://docs.legis.wisconsin.gov/misc/lfb/informational_papers/january_2017/0018_shared_revenue_program_informational_paper_18.pdf) for calculating the level of shared revenue, **population largely determines the level of shared revenue that each county receives**.[^2] The graph below shows total shared revenues for each county against county population (both shown on a log-scales for readability). Counties in which Trump won a majority of votes in the 2020 general election are shown in red (those where he did not get a majority in blue). Counties above the line were given higher per-capita shared revenue allocations than average. Notably both Dane and Milwaukee counties were given above average per-capita allocations (the two blue dots in the upper right of the graph).

[^2]: Every person in a county's population earns it roughly an additional \$5k in shared revenue, determined by a linear regression model R^2^ = 0.93.

```{r, population_graph}
#| echo: false
#| warning: false


ggplot(data = data, aes(x = log(Population), y = log(`Total Shared Revenue`))) + 
            geom_point(aes(color = `2020 Trump Majority`)) +
  scale_colour_manual(values = c("blue", "red")) + 
  geom_abline(slope = slope, intercept = intercept) + 
  labs(title = "WI County (and Municipal) Shared Revenue by Population",
       x = "County Population (log)", y = "Tota Shared Revenue in $ (log)") +
theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x =element_blank() , 
        axis.text.y = element_blank() , 
        axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank())



```

Although *individuals* are given roughly equal allocations regardless of where they live, this can result in local governments receiving proportionally less money in shared revenue than their residents pay in taxes. Richer counties tend to receive less than they pay in, and poorer counties more, with total tax revenue serving as a reasonable proxy for per-capita wealth. The graph below and to the left shows the amount of shared revenue counties are allocated above their relative (per capita) contribution through income and sales taxes. This is the amount they are subsidized by tax payers in other counties, with bars going to the left showing a *negative* subsidy. Counties are ordered (in descending order) by per-capita tax revenue, showing that poorer counties generally receive larger subsidies.[^3] It also shows that there is a great deal of variability in subsidy between counties of similar per-capita tax.

[^3]: By log-linear regression model, R^2^ = 0.59.

The graph below and to the right shows the percentage change in per-capita shared revenue if there were no subsidy, meaning that each county received shared revenue proportional to their total tax contribution. Because of the patterns of wealth in Wisconsin, an end (or scaling back) of the shared governance subsidy would be particular devastating for primarily rural poorer counties plus Milwaukee county while benefiting rich primarily suburban ones.

::: {layout-ncol="2"}
```{r, shared_revenue_subsidy, fig.height = 10 , fig.width = 6}
#| echo: false
#| warning: false

ggplot(data = data, aes(y = reorder(County,`Per capita Combined Tax`) , x = `Per capita Shared Revenue Subsidy`)) +
  scale_fill_manual(values = c("blue", "red")) + 
  geom_bar(aes(fill = `2020 Trump Majority`), stat = "identity") +
  theme_minimal() +
labs(title = "Per Capita Shared Revenue Subsidy",
     subtitle = "Ordered by per-capita tax procedes (highest to lowest)",
       x = "Per Capita $", y = "") +     
  theme(legend.position="bottom")
  
    
```

```{r, Change_InFunding, fig.height = 10 , fig.width = 6}
#| echo: false
#| warning: false

ggplot(data = data, aes(y = reorder(County,`Per capita Combined Tax`) , x = ( (`Total Shared Revenue` - `Total Shared Revenue Subsidy`) - `Total Shared Revenue`)/`Total Shared Revenue`)) +
  scale_fill_manual(values = c("blue", "red")) + 
   scale_x_continuous(labels = scales::percent) +
  geom_bar(aes(fill = `2020 Trump Majority`), stat = "identity") +
  theme_minimal() +
labs(title = "% Change in Shrd. Rev. if Based on Taxes Only",
     subtitle = "Ordered by per-capita tax procedes (highest to lowest)",
       x = "% Change in Shared Revenue", y = "") +     
  theme(legend.position="bottom")
  
    
```
:::

Although per-capita shared revenue is largely similar across counties, rural counties earn higher per-capita shared revenue even when accounting for difference levels of wealth (as represented by total tax proceeds).[^4] The below graph shows these results: per-capita shared revenue against population density, with each quartile of per-capita tax proceeds represented by a different pane.

[^4]: By Log-linear regression model, R^2^ = 0.53. Model coefficients are no longer significantly different from 0 when including an interaction effect between density and per-capita tax proceeds. This may be due to there being too few data points (counties) to adequately model the interaction, but might be an area of further investigation (by someone with more knowledge of statistics).

```{r, density}




#linear model of density and per capita shared revenue, with individual model per quartile.
density.percapita.combinedtax.quartile1 <-  data %>% 
  mutate(`Tax Quartile` = ntile(`Per capita Combined Tax`, 4)) %>%
  filter(`Tax Quartile` == 1) %>%
lm(log((`Total Shared Revenue`)/Population) ~ log(`Density (pop sq mi)`), data = .)
#summary(density.percapita.combinedtax.quartile1)

intercept.slope.df <- data.frame(quartile = 1, 
                                 intercept = as.numeric(coef(density.percapita.combinedtax.quartile1)[1]),
                                 slope = as.numeric(coef(density.percapita.combinedtax.quartile1)[2]))

intercept.1 <- as.numeric(coef(density.percapita.combinedtax.quartile1)[1])
slope.1 <- as.numeric(coef(density.percapita.combinedtax.quartile1)[2])

density.percapita.combinedtax.quartile2 <-  data %>% 
  mutate(`Tax Quartile` = ntile(`Per capita Combined Tax`, 4)) %>%
  filter(`Tax Quartile` == 2) %>%
  lm(log((`Total Shared Revenue`)/Population) ~ log(`Density (pop sq mi)`), data = .)
#summary(density.percapita.combinedtax.quartile2)

intercept.slope.df <- rbind(intercept.slope.df,data.frame(quartile = 2, 
                                   intercept = as.numeric(coef(density.percapita.combinedtax.quartile2)[1]),
                                   slope = as.numeric(coef(density.percapita.combinedtax.quartile2)[2])))

intercept.2 <- coef(density.percapita.combinedtax.quartile2)[1]
slope.2 <- coef(density.percapita.combinedtax.quartile2)[2]

density.percapita.combinedtax.quartile3 <-  data %>% 
  mutate(`Tax Quartile` = ntile(`Per capita Combined Tax`, 4)) %>%
  filter(`Tax Quartile` == 3) %>%
  lm(log((`Total Shared Revenue`)/Population) ~ log(`Density (pop sq mi)`), data = .)
#summary(density.percapita.combinedtax.quartile3)


intercept.slope.df <- rbind(intercept.slope.df,data.frame(quartile = 3, 
                                    intercept = as.numeric(coef(density.percapita.combinedtax.quartile3)[1]),
                                     slope = as.numeric(coef(density.percapita.combinedtax.quartile3)[2])))

intercept.3 <- coef(density.percapita.combinedtax.quartile3)[1]
slope.3 <- coef(density.percapita.combinedtax.quartile3)[2]

density.percapita.combinedtax.quartile4 <-  data %>% 
  mutate(`Tax Quartile` = ntile(`Per capita Combined Tax`, 4)) %>%
  filter(`Tax Quartile` == 4) %>%
  lm(log((`Total Shared Revenue`)/Population) ~ log(`Density (pop sq mi)`), data = .)
#summary(density.percapita.combinedtax.quartile4)


intercept.slope.df <- rbind(intercept.slope.df,data.frame(quartile = 4, 
                                  intercept = as.numeric(coef(density.percapita.combinedtax.quartile4)[1]),
                                slope = as.numeric(coef(density.percapita.combinedtax.quartile4)[2])))
intercept.4 <- coef(density.percapita.combinedtax.quartile4)[1]
slope.4 <- coef(density.percapita.combinedtax.quartile4)[2]

data %>% mutate(quartile = ntile(`Per capita Combined Tax`, 4)) %>%
inner_join(intercept.slope.df, by = "quartile") %>%
mutate(quartile.text = ifelse(quartile > 1, "25% Below Median Tax","Bottom 25% Tax"),
       quartile.text = ifelse(quartile > 2, "25% Above Median Tax", quartile.text),
       quartile.text = ifelse(quartile > 3, "Top 25% Tax", quartile.text),
       quartile.text = factor(quartile.text, levels = c("Bottom 25% Tax","25% Below Median Tax",
                                                        "25% Above Median Tax", "Top 25% Tax"))) %>%
ggplot(data = ., aes(x = log(`Density (pop sq mi)`), y = log((`Total Shared Revenue`)/Population))) +
  geom_point(aes(color = `2020 Trump Majority`)) +
  scale_color_manual(values = c("blue", "red")) + 
  theme_minimal() +
labs(title = "Per Capita Shared Revenue by Density and Per Capita Tax Proceeds (Panes)",
       x = "Population Density (log)", y = "Per Capita Shared Revenue (log)") +     
  theme(legend.position="bottom")+
  facet_grid(cols = vars(quartile.text)) + 
  geom_abline(aes(slope = slope, intercept = intercept))  +
    theme(axis.text.x = element_blank(),
        axis.ticks.x =element_blank() , 
        axis.text.y = element_blank() , 
        axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank())


```

## Is Wisconsin's Shared Revenue System Fair?

There are strong norms of fairness that require citizens to have equal access to government services regardless of wealth. For instance, most people would consider it unjust for a government to provide different tiers of services based upon the amount an individual pays to the government in taxes. Much the same can be said for providing larger shares of shared revenue to some *communities* based upon the higher contribution of their members of tax revenues. In this respect, the current method that provides all communities with similar per-capita shared revenue allocations is fair.

An admittedly more speculative argument might be made to justify higher per-capita shared revenue allocations to less dense areas or areas with large amounts of older infrastructure (like Milwaukee county). Because services are more expensive to provide in these areas, and what matters to people is the quality of services provided rather than how much money it takes to provide them, it may be fairer to provide funds necessary for equality of service quality rather than equal funds. Of course, the more expensive cost to provide services in rural areas may just as easily justify the government in providing incentives to abandon rural communities where jobs are scarce and the provision of services is more expensive.

Even without evaluating these arguments it is easy to see that much of the dominant narratives about shared revenue aren't based in fact. For instance, the commonly expressed idea that hard-working conservative people in rural and suburban counties subsidize the profligacy of liberals in Madison and Milwaukee is clearly false. Most rural counties are heavily subsidized by tax-payers in more urban ones, including Dane county, and the per-capita subsidy provided to Milwaukee county is much smaller than most rural counties.

Some have also claimed that the shared revenue system is explicitly unfair to more liberal communities, like Madison and Milwaukee. However, the political affiliation of a county doesn't appear to have any impact on its shared revenue allocation. Dane and Milwaukee counties receive per-capita allocations slightly above what would be expected from their population alone. And although Dane county does have a substantial negative subsidy, so do other more suburban counties that are much more conservative.

I think any substantial changes to shared revenue allocation - like allocating based upon tax revenues rather than population - would require a deeper discussion of an alternative notion of fairness than the one I describe above. There is a model for this in how local property taxes are largely collected by and allocated to the communities in which the are raised. Perhaps it can be argued that tax payers have a substantial interest in having their income and sales taxes stay in their communities where it will impact them more and be easier for them to control (through local government). However, allowing richer communities to keep more taxes of their residents would have a devastating impact on services in poorer communities, including Milwaukee county. Additionally, it seems to ignore existing criticisms of how property taxes are currently kept by the communities in which they are raised, especially how this system leads to under-funding of schools and various social services in poorer communities. It would be odd for liberals to argue for its extension to all forms of taxation.
